The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-04-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions on section 2, notation of goodness-of-fit measure and other small corrections (including typos in the references)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2524 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3115ec7f726a60cbd05c39501ecca12baee96081

if(file.exists("stump.rda")){
  load("stump.rda")
} else {
  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
                         nrsplitvar = 1, nrsteps = 1, nrlevels = 2,
                         delta = seq(from = 0, to = 1, by = 0.1),
                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = c("all", "beta0", "beta1"),
                         binary_regressor = FALSE, binary_beta = TRUE, only_intercept = FALSE,
                         test = c("ctree", "mfluc", "ctree_max",
                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
                         beta0 = 0, beta1 = 1,
                         stump = TRUE, z1dist = "unif", sigma = 1, alpha = 0.05)

@@

Therefore, three different popular approaches are considered here: classical categorical 
association tests (GUIDE), conditional inference (CTree), and parameter instability 
tests (MOB). First, these are embedded into a common inference framework encompassing
parametric model trees, in particular linear model trees. Second, it is assessed how
different building blocks from this common framework affect the power of the 
tree algorithms to select the appropriate covariates for splitting: observation-wise
goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
at zero, binning of covariates, conditional vs.\ unconditional approximations of the
null distribution. This shows that specifically the goodness-of-fit measure is crucial
for the power of the procedures, with model scores without binarization performing
much better in many scenarios.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Transformations of a goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Binarization of these residuals or scores.
(3)~Categorization of covariates.
(4)~Approximations of the null distribution: conditional vs.\ unconditional.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables clear recommendations can be given based on the presented 
results.

@@

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various different tree 
algorithms have been developed, all following the same general structure starting at 
the root of the tree which is simply the first node containing all available data:
\begin{enumerate}
\item For the data set in the current node some form of discrepancy 
measure is calculated. 
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached. 
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a parametric model fitted in each 
node. For instance, different types of residuals or their signs can be applied 
\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores as 
in \cite{Hothorn+Hornik+Wiel:2006}.
%(In the simulation study of this paper trees with linear models fitted in each node are 
%investigated with the residual sum of squares (RSS) employed as discrepancy measure.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then, in a separate step, the split point. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item A discrepancy measure is obtained in the form of an $n \times k$-matrix $D$ giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and the parameter $k \in \mathbb{N}$ defined by the algorithm.
\item Using this matrix the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $D$ on $Z_j$.
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide} .

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
$$
s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to obtain the estimated model parameter $\hat{\beta}$. For this purpose many different M-type 
estimators can be applied including maximum likelihood or ordinary least squares such that the 
corresponding objective function can be employed as model-based discrepancy measure for the 
tree algorithm. Evaluating the score function $s$ on a data set of $n$ observations 
%containing the 
%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
and a $k$-dimensional estimated parameter vector 
%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

While CTree offers to apply model-based discrepancy measures as one possible choice, the 
MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
parametric models into the framework of regression trees and is hence purely intended for 
a model-based approach.

@@

Before applying a statistical test two additional steps are performed:
(1)~The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$.  % FIX ME: +1 and -1 in the implementation? 
(2)~For each possible split variable the observed values are categorized into four groups 
separated by the corresponding 1st, 2nd and 3rd quartile unless the split variable is already
categorical.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables and is in that way fixed.

\newpage
\subsection{Pruning}
\label{sec:pruning}
In early developed methods of unbiased recursive partitioning such as 
FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
the $p$-values of applied significance tests are used purely for the selection of a split variable 
based on a unified scale. All other steps are performed in the same way as in classic tree 
algorithms, such as finding the best split point by optimizing an objective function.

@@

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
cases.
By including all available scores rather than just residuals the goodness of fit of the model is 
investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

\newpage

@@

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e. only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
only scores has already been discussed and elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992}.

@@

\subsection{Full factorial analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying and the true split point is in the
center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
$\delta = 0.3$ is considered making the differences particularly well observable 
(additional easing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.65),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
                                              expression(xi==0.8~"and"~delta==1)),
                                   bg = "gray90"),
       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
                  lines = list(lty = c(1,1,2,2), 
                               type = "l", lwd = 2,
                               col = c(pal[1], pal[3], pal[1], pal[3]))))

>>>>

if(file.exists("stump.rda")){
  load("stump.rda")
} else {
  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
                         nrsplitvar = 1, 
                         nrsteps = 1, nrlevels = 2,
                         delta = seq(from = 0, to = 1, by = 0.1),
                         xi = c(0, 0.2, 0.5, 0.8), 
                         vary_beta = c("all", "beta0", "beta1"),
                         binary_regressor = FALSE,
                         binary_beta = c(TRUE, FALSE),
                         only_intercept = FALSE,
                         test = c("ctree", "mfluc", "ctree_max",
                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
                       beta0 = 0, beta1 = 1,
                       stump = TRUE, z1dist = "unif", sigma = 1, alpha = 0.05)

@@

Therefore, three different popular approaches are considered here: classical categorical 
association tests (GUIDE), conditional inference (CTree), and parameter instability 
tests (MOB). First, these are embedded into a common inference framework encompassing
parametric model trees, in particular linear model trees. Second, it is assessed how
different building blocks from this common framework affect the power of the 
tree algorithms to select the appropriate covariates for splitting: observation-wise
goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
at zero, binning of covariates. This shows that specifically the goodness-of-fit measure 
is crucial for the power of the procedures, with model scores without binarization performing
much better in many scenarios.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Variations of a goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Binarization of these residuals or scores.
(3)~Categorization of covariates.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

In many of the considered scenarios it can be observed that the choice of goodness-of-fit
measure heavily influences the performance of testing strategies as overall using model 
scores leads to clearly better results than employing residuals only. Moreover, the original
values of the goodness-of-fit measure are to be preferred over binarized versions of them.
Also regarding the effects of categorizing possible split variables and the selection
of a pruning strategy clear recommendations can be given based on the presented results.

@@

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various different tree 
algorithms have been developed, all following the same general structure starting at 
the root of the tree which is simply the first node containing all available data:
\begin{enumerate}
\item For the data set in the current node a model is fitted by optimizing
some objective function that reflects the goodness of fit.
\item Among all available covariates one is selected as the split variable,
choosing the split point such that goodness of fit is maximized in the
resulting subgroups.
\item Steps 1 and 2 are repeated within each subgroup until some stopping
criterion is attained.
\end{enumerate}

The term ``model'' is used here in a very broad sense and encompasses not
only least-squares or maximum-likelihood models but also simple constant
fits such as means or average proportions.
Therefore, depending on the type of model, the employed objective function can 
for example be the sum of deviations from a typical or average value, but it can 
also be constructed based on the model fitted in each node and depend not only on 
the response but also on possible covariates. For instance, different 
types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
well as rank sums or logrank scores as in \cite{Hothorn+Hornik+VanDeWiel:2006}.

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, 
MOB, and GUIDE first select the split variable and then, in a separate step, the 
split point. For the first step of selecting a split variable they all apply statistical 
tests following the same general strategy: 
\begin{enumerate}
\item To capture how the objective function changes with the observations in
the current subgroup, a disaggregated, observation-wise goodness-of-fit
measure is obtained. More formally, this is an $n \times k$ matrix $G$
where $n \in \mathbb{N}$ is the number of observations and $k \in \mathbb{N}$ 
the number of measures per observation.
\item The dependency or association of this goodness-of-fit matrix $G$ with
each possible split variable $Z_j$, $j \in \{1, \dots, m\}$, is assessed
using some suitable test statistic. The corresponding $p$-values allow for
a comparison of all $m$ covariates on a standardized or unified scale and in
that way for an unbiased split variable selection.
\item The covariate corresponding to the smallest $p$-value -- and thus the
highest influence on the goodness of fit of the model -- is selected for 
splitting the data into subgroups.
\end{enumerate}

As explained in more detail below, the aggregated overall objective
function in the linear regression model would be the residual sum of
squares and the corresponding observation-wise goodness-of-fit measure is
given by the residuals or scores (gradient contributions). Analogously,
the log-likelihood and corresponding score function could be used.

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
$$
s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the aggregated overall objective 
function that is optimized to obtain the estimated model parameter $\hat{\beta}$. 
For this purpose many different M-type estimators can be applied including maximum 
likelihood or ordinary least squares such that the corresponding objective functions 
can also be employed for the tree algorithm. Evaluating the score function $s$ on a 
data set of $n$ observations 
%containing the 
%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
and a $k$-dimensional estimated parameter vector 
%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

While CTree offers to apply model-based goodness-of-fit measures as one possible choice, the 
MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
parametric models into the framework of regression trees and is hence purely intended for 
a model-based approach.

@@

Before applying a statistical test two additional steps are performed:
(1)~The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$.  % FIX ME: +1 and -1 in the implementation? 
(2)~For each possible split variable the observed values are categorized into four groups 
separated by the corresponding 1st, 2nd and 3rd quartile unless the split variable is already
categorical.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
The possible number of breakpoints is determined by the categorization of the possible 
split variables and is in that way fixed.

%\newpage
\subsection{Pruning}
\label{sec:pruning}
In early developed methods of unbiased recursive partitioning such as 
FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
the $p$-values of applied significance tests are used purely for the selection of a split variable 
based on a unified scale. All other steps are performed in the same way as in classic tree 
algorithms, such as finding the best split point by optimizing an overall goodness-of-fit measure.

@@

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item Variations of the goodness-of-fit measure:\\
Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
model that is obtained for the response $Y$ and possible covariates $X$ can be 
considered. This can be residuals $r(Y,X,\hat{\beta})$ as in the GUIDE algorithm or 
scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
transformation function $h$ as in the CTree algorithm including both of the foregoing cases.
By including all available scores rather than just residuals the goodness of fit of the 
model is investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

A very basic form of goodness-of-fit measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach one can consider residuals defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding disaggregated, observation-wise goodness-of-fit measure is defined by
the score function 
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e. only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
only scores has already been elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
in Section~\ref{sec:discussion}.

@@

\subsection{Full factorial analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e. $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
in order to make the differences particularly well observable.
In the right panel the true split point is close to the upper end of the considered interval
($\xi=0.8$) requiring a higher effect size $\delta = 1$ in order to visualize results properly.
(Additional evaluations for increasing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
                                              expression(xi==0.8~"and"~delta==1)),
                                   bg = "gray90"),
       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
                  lines = list(lty = c(1,1,2,2), 
                               type = "l", lwd = 2,
                               col = c(pal[1], pal[3], pal[1], pal[3]))))
